Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)

### 2018-02-25 paper.tex
update

%% Title information
\title[Short Title]{Pushdown Control-Flow Analysis via Refunctionalization}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Filling the Gap between Abstracting Abstract Machine and Abstracting Definitional Interpreter}        %% \subtitle is optional
\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

\item We present an abstract interpreter obtained from refunctionalizing
  an small-step abstracting abstract machine. 
  This abstract interpreter naturally has pushdown control-flow property
  since we represent the evluation context of analyzed language using
  evluation context of the meta-language.
  We also analyze its termination, soundness and complexity.

@@

We use Scala language to demonstrate the idea and each step of transformations.

@@

\subsection{CESK Machine}

@@

\subsection{Abstracting Abstract Machine} \label{aam}
Abstracting abstract machine is a systematic methodology that derives sound abstract interpreters
from a concrete abstract machine \cite{van2012systematic, van2010abstracting}.

There are several notable differences of store between AAM and concrete CESK machine.
In AAM, the store maps addresses to sets of values, so it stores all possible values
for a particular address. Thus the dereference an address becomes non-deterministic.
Also, the store performs a join rather than overwriting when updating elements of the store.
Third, the continuations are allocated on store instead of formed a linked list.

For clarity, we divide the store into binding store \texttt{BStore} and continuation store \texttt{KStore}. 
The binding store maps addresses of binding to sets of closure values, and continuation store maps
continuation addresses to sets of continuations.
We define a generic class \texttt{Store[K,V]} that performs join when updating elements
in a store. By parametering \texttt{Store[K,V]} with \texttt{[BAddr, Storable]} and 
\texttt{[KAddr, Cont]}, we obtain \texttt{BStore} and \texttt{KStore}, perspectively.

@@

\begin{verbatim}
case class Store[K,V](map: Map[K, Set[V]]) {
  def apply(addr: K): List[V] = map(addr).toList
  def update(addr: K, d: Set[V]): Store[K,V] = {
    val oldd = map.getOrElse(addr, Set())
    Store[K, V](map ++ Map(addr -> (d ++ oldd)))
  }
  def update(addr: K, sd: V): Store[K,V] = update(addr, Set(sd))
}

type BStore = Store[BAddr, Storable]
type KStore = Store[KAddr, Cont]
\end{verbatim}

As a consequence, the components of a state is also changed: 
the store is divided into binding store and continuation store;
the continuation stack becomes an address that maps to a set of continuation.
By dereferencing this address in continuation store, we can retrieve the 
actually transfer of controls.

@@

The definitions of closure \texttt{Clos}, stack frame \texttt{Frame}, and 
environment \texttt{Env} remain the same as concrete CESK machine.

\begin{verbatim}
case class BAddr(x: String, time: Time)
type Env = Map[String, BAddr]

abstract class KAddr
case object Halt extends KAddr
case class ContAddr(tgt: Expr, time: Time) extends KAddr

@@

\begin{verbatim}
def allocBind(x: String, time: Time): BAddr = BAddr(x, time)
def allocKont(tgtExpr: Expr, tgtEnv: Env, tgtStore: BStore, time: Time): KAddr = 
  ContAddr(tgtExpr, time)
type Time = List[Expr]
def k: Int = 0
def tick(s: State): Time = (s.e::s.time).take(k)

def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())

@@

\begin{verbatim}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, kstore, kaddr, time) 
      if isAtomic(ae) =>
      val closures = atomicEval(f, env, bstore).toList
      for (Clos(Lam(v, body), env_c) <- closures) yield {
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val newBStore = bstore.update(baddr, atomicEval(ae, env, bstore))
        val newKAddr = allocKont(body, env_c, newBStore, new_time)
        val newKStore = kstore.update(newKAddr, Cont(Frame(x, e, env), kaddr))
        State(body, new_env, newBStore, newKStore, newKAddr, new_time)
      }
    case State(ae, env, bstore, kstore, kaddr, time) 
      if isAtomic(ae) =>
      for (Cont(Frame(x, e, f_env), f_kaddr) <- kstore(kaddr).toList) yield {
        val baddr = allocBind(x, new_time)
        val new_env = f_env + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(e, new_env, new_store, kstore, f_kaddr, new_time)
      }
  }
}
\end{verbatim}

@@

\begin{verbatim}
def drive(todo: List[State], seen: Set[State]): Set[State] = todo match {
  case Nil => seen
  case hd::tl if seen.contains(hd) => drive(tl, seen)
  case hd::tl => drive(step(hd).toList ++ tl, seen + hd)
}

@@

def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) =>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, new_env, new_store, frame::konts, new_time)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, f_env)::konts =>
          val baddr = allocBind(x, new_time)
          val new_env = f_env + (x -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, new_env, new_store, konts, new_time))
      }
  }
}
\end{verbatim}

@@

Recent years, there are great efforts \cite{vardoulakis2010cfa2,
  earl2012introspective, gilray2016pushdown}
to achieve precise call/return match based
on small-step abstracting abstract machines which is we described in section \ref{aam}.

@@

\section{FROM AAM TO ADI}

>>>>

%% Title information
\title[Short Title]{Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Filling the Gap between Abstracting Abstract Machine and Abstracting Definitional Interpreter}        %% \subtitle is optional
\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

\item We present an abstract interpreter obtained from refunctionalizing
  an small-step abstracting abstract machine. 
  This abstract interpreter naturally has pushdown control-flow property
  since we represent the evluation contexts of analyzed language using
  continuations of the meta-language, and therefore sequentialize the
  analyzing.
  We also analyze its termination, soundness and complexity.

@@

We use Scala language to demonstrate the idea and each step of transformations.
There are two main reasons to do so:
1) The code does not diminish the accuracy of the material
than formal and mathematical notations, 
which are heavily used in other papers in static analysis or semantics.
The Scala code in this paper can be easily back-translated into formal notations.
2) The code is also executable with only few changes, which make it particularly fit for 
presenting topic: syntactical transformations of programs.

@@

\subsection{CESK Machine} \label{cesk}

@@

\subsection{Abstracting Abstract Machine} \label{aam}
Abstracting abstract machine (AAM) is a systematic methodology that derives sound 
abstract interpreters for higher-order functional languages from concrete 
abstract machines \cite{van2012systematic, van2010abstracting}. 
An abstracting abstract machine implements computable abstract semantics that 
approximates the runtime behaviors of programs.
Since the state space of concrete execution is possibily infinite,
so to analyze programs, the key insight of AAM approach is to allocate both bindings
and continuations on the store, and then bound the addresses space to be finite.
Thus the approximated machine-state space is finite, and computable.

In this section, we derive the abstracting abstract machine from concrete
CESK machines, and also show that how to instantiate
useful $k$-call-sensitive analysis.

Similar with CESK machines, the machine state of the AAM has a control expression,
an environment, a store and continuation, and in addition with a time-stamp.
But there are several notable differences of store between AAMs and concrete CESK machines.
In AAM, the store maps addresses to sets of values, so it stores all possible values
for a particular address. Thus the dereference an address becomes non-deterministic.
Also, the store performs a join rather than overwriting when updating elements of the store.
Third, the continuations are likewise allocated on store instead of formed a linked list,
and the continuation in a state then turns into a continuation address.

For clarity, we divide the store into binding store \texttt{BStore} and continuation store \texttt{KStore}. 
The binding store maps binding addresses to sets of closure values, and continuation store maps
continuation addresses to sets of continuations.
We define a generic class \texttt{Store[K,V]} that performs join when updating elements
in a store. By parameterizing \texttt{Store[K,V]} with \texttt{[BAddr, Storable]} and 
\texttt{[KAddr, Cont]}, we obtain \texttt{BStore} and \texttt{KStore}, perspectively.

@@

\begin{verbatim}
case class Store[K,V](map: Map[K, Set[V]]) {
  def apply(addr: K): List[V] = map(addr).toList
  def update(addr: K, d: Set[V]): Store[K,V] = {
    val oldd = map.getOrElse(addr, Set())
    Store[K, V](map ++ Map(addr -> (d ++ oldd)))
  }
  def update(addr: K, sd: V): Store[K,V] = update(addr, Set(sd))
}
type BStore = Store[BAddr, Storable]
type KStore = Store[KAddr, Cont]
\end{verbatim}

The co-domain of binding stores is \texttt{Storable} which is the same
as we defined for CESK machines.
The co-domain of continuation stores \texttt{Cont} is comprised of 
a frame \texttt{Frame} and a continuation address \texttt{KAddr}.
To mimic the runtime call stack, \texttt{KAddr} plays the role of
representing the remaining stack frames.
TODO: nondet of cont?

\begin{verbatim}
abstract class Storable 
case class Clos(v: Lam, env: Env) extends Storable

As a consequence, the components of a state is also changed: 
the store is divided into binding store and continuation store;
the continuation stack becomes an address.
By dereferencing this address in continuation store, we can retrieve the 
actual transfer of controls. The definition of environment \texttt{Env}
remains the same.

@@

Up to now, we did not describe much about the time-stamp \texttt{Time}
os states and addresses of stores.
But they are key ingredients to achieve analysis with different sensitivities,
as well as a finite state space analysis.
To effectively approximate the runtime behavior, a finite 
program contour \texttt{time} that encodes the program execution history is introduced.
Here, we use a list of execution contexts (expressions) to reprent that, and
as we will see in section \ref{kcfainst}, by using different \texttt{tick}
function on time we may obtain a family of analyses.

\begin{verbatim}
type Time = List[Expr]
\end{verbatim}

Address.

\begin{verbatim}
case class BAddr(x: String, time: Time)

abstract class KAddr
case object Halt extends KAddr
case class ContAddr(tgt: Expr, time: Time) extends KAddr
\end{verbatim}

\subsubsection{Single-step Evaluation}

@@

\begin{verbatim}
def allocBind(x: String, time: Time): BAddr = BAddr(x, time)
def allocKont(tgtExpr: Expr, tgtEnv: Env, 
              tgtStore: BStore, time: Time): KAddr = ContAddr(tgtExpr, time)

@@

\begin{verbatim}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, kstore, kaddr, time) 
      if isAtomic(ae) =>
      val closures = atomicEval(f, env, bstore).toList
      for (Clos(Lam(v, body), env_c) <- closures) yield {
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_bstore = bstore.update(baddr, atomicEval(ae, env, bstore))
        val new_kaddr = allocKont(body, env_c, new_bstore, new_time)
        val new_kstore = kstore.update(new_kaddr, Cont(Frame(x, e, env), kaddr))
        State(body, new_env, new_bstore, new_kstore, new_kaddr, new_time)
      }
    case State(ae, env, bstore, kstore, kaddr, time) 
      if isAtomic(ae) =>
      val conts = kstore(kaddr).toList
      for (Cont(Frame(x, e, env_f), f_kaddr) <- conts) yield {
        val baddr = allocBind(x, new_time)
        val new_env = env_f + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(e, new_env, new_store, kstore, f_kaddr, new_time)
      }
  }
}
\end{verbatim}

\subsubsection{$k$-call-sensitive Instantiation} \label{kcfainst}

\begin{verbatim}
def k: Int = 0
def tick(s: State): Time = (s.e::s.time).take(k)
\end{verbatim}

\subsubsection{Collecting Semantics}

@@

\begin{verbatim}
def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())

def drive(todo: List[State], seen: Set[State]): Set[State] = todo match {
  case Nil => seen
  case hd::tl if seen.contains(hd) => drive(tl, seen)
  case hd::tl => drive(step(hd).toList ++ tl, seen + hd)
}

@@

def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) if isAtomic(ae) =>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, new_env, new_store, frame::konts, new_time)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, env_f)::konts =>
          val baddr = allocBind(x, new_time)
          val new_env = env_f + (x -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, new_env, new_store, konts, new_time))
      }
  }
}
\end{verbatim}

@@

Recent years, there are great efforts \cite{vardoulakis2010cfa2, earl2012introspective, gilray2016pushdown}
to achieve precise call/return match based
on small-step abstracting abstract machines which is what we described in section \ref{aam}.

@@

\section{From AAM to ADI}
